---
title: "STA141A Final"
subtitle: "STA141A | Fall 2022 | Kühnert"
author: "Alvin Akpokli, Sarvesh Krishan, Ngai Pan Ng, Hannah Spray"
date: "7 December 2022"
output: 
  pdf_document:
    keep_md: true
fontsize: 10pt
---

## Project Overview  
This project aims to investigate possible risk factors for death due to heart failure.  

### Group members:  
- Alvin Akpokli (akakpokli@ucdavis.edu)
- Sarvesh Krishan (skrishan@ucdavis.edu)
- Ngai Pan Ng (npng@ucdavis.edu)
- Hannah Spray (hgspray@ucdavis.edu)

### Research questions of interest:  
1. Is death from heart failure significantly more prevalent in those with high blood pressure (i.e. hypertension)?
2.  Is death from heart failure significantly more prevalent in men or women that smoke versus those who do not?

### Dataset:  
Our dataset is called "Heart failure clinical records dataset". This dataset was obtained from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Heart+failure+clinical+records).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadData, echo = FALSE}
# make sure csv file is in working directory
data = read.csv("heart_failure_clinical_records_dataset.csv")
head(data)
length(data$DEATH_EVENT)
names(data)

y = data$DEATH_EVENT

```


\pagebreak
## References  
Davide Chicco, Giuseppe Jurman: "Machine learning can predict survival of patients with heart failure from serum creatinine and ejection fraction alone". BMC Medical Informatics and Decision Making 20, 16 (2020).
